Using Information Underlying Missing Data to Improve Estimation of NFL Field Goal Kicker Accuracy

### 2016-11-19 JaradNiemi_MwSAM2016.Rnw
Add slide version 1
Add version 2

\documentclass{beamer}

@@

Let's compare two field goal kickers and their proportion of field goals made 
for the 2000-2011 seasons:

@@

\frametitle{Raw statistics}
<<raw_statistics, results="asis">>=
kicker_names <- NFLKicker::kicker_names
kicker_names$kicker_hidden <- c("Kicker 2","Kicker 1")

fg <- NFLKicker::FGdata %>%
  right_join(kicker_names, by = "fkicker")

fg_raw <- fg %>%
  group_by(kicker_hidden) %>%
  summarize(n = length(good),
            made = sum(good)) %>%
  mutate(p = made/n)

@@

<<proportion_table, results="asis">>=
fg_table %>%
  dplyr::select(`Kicker`, `Proportion made`) %>%
  xtable(digits = c(1,NA,2)) %>%
  print(include.rownames = FALSE)
@

@@

\begin{frame}
\frametitle{Posterior distribution for true probability}

<<posterior, dependson="raw_statistics", fig.width=8>>=
create_posterior <- function(d) {
  data.frame(probability = seq(0.6,1, length=101)) %>%
    mutate(density = dbeta(probability, d$made+.5, d$n-d$made+.5))
}

posterior <- fg_raw %>%
  group_by(kicker_hidden) %>%
  do(create_posterior(.))

ggplot(posterior, aes(probability, density, color=kicker_hidden, group=kicker_hidden)) +
  geom_line() +
  theme_bw() + 
  labs("Probability of making a field goal", ylab="Posterior density", color="Kicker")
@
\end{frame}

@@

<<fg_dist, fig.width=10, out.width="\\textwidth">>=
fg <- NFLKicker::FGdata %>%
  right_join(kicker_names, by = "fkicker")

fg_dist <- fg %>%
  group_by(kicker_hidden, dist) %>%
  summarize(n = length(good),
            made = sum(good)) %>%
  mutate(p = made/n)

@@

<<results='asis'>>=
fg_probit <- fg %>%
  mutate(Surface       = ifelse(surf == "Grass", "Grass", "Synthetic"))

@@

<<glm, fig.width=10, out.width="\\textwidth">>=
glm_predictions <- function(d) {
  m <- glm(good ~ dist+Surface, data=d, family="binomial")
  dd <- expand.grid(dist=17:80, Surface=c("Grass","Synthetic"))
  dd$probability <- as.numeric(predict(m, dd, type="response"))
  return(dd)
}

@@

\begin{frame}
\frametitle{Hierarchical probit regression model}
<<hierarchical_model, warning = FALSE>>=
fg <- NFLKicker::FGdata %>%
  mutate(Surface       = ifelse(surf == "Grass", "Grass", "Synthetic"),
         scaled_dist = as.numeric(scale(dist, center=36, scale = 10)))

m_glmer <- glmer(good ~ (Surface*scaled_dist|fkicker), 
           data = fg, 
           family = binomial(link="probit"))
@

@@

For opportunity $i$ for kicker $k$, let 
\begin{itemize}
\item $A_{ik}$ be an indicator of attempt, i.e. 1 if a kick was attempted and 0 otherwise, and 
\item $w_{ik}$ be a vector of explanatory variables.
\end{itemize}

@@

The full model is 
\[ \begin{array}{rlrlrl}
Y_{ak} &\ind Ber(\theta_{ak}) & \theta_{ak}  &= \Phi\left(     x_{ak}^\top \beta_k \right) & \beta_k \ind N(\mu,\Sigma) \\
A_{ik} &\ind Ber(\pi_{ik})    & \pi_{ik}     &= \Phi\left(\omega_{ik}^\top \alpha\phantom{_k}  \right) \\
\omega_{ik} &= (w_{ik}, w_{ik}\Phi^{-1}(\theta_{ik}))
\end{array} \]
\pause
where 
\begin{itemize}
\item $x_{ak}$ is a set of explanatory variables that affect the probability of making a kick \pause and
\item $w_{ik}$ is a set of explanatory variables that affect the probability of taking a kick. 
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Informative missingness analysis}
<<informative_missing_data>>=
library(dplyr)

@@

<<informative_missing_plot, dependson="informative_missing_data", fig.width=10, out.width="\\textwidth">>=
ggplot(p_d, aes(dist, probability, color=Surface)) +
#  geom_point(data=fg_dist2, aes(dist, p, size=n)) +
  geom_line(aes(group=Surface)) +
  facet_grid(kicker_hidden ~.) +
  theme_bw() +
  labs(x="Distance (yards)",y="Probability of making the kick", size="Number of\nattempts") +
  ylim(0,1) +
  scale_color_manual(values = c("Grass" = "seagreen", "Synthetic" = "Black"))
@
\end{frame}

@@

\begin{frame}
\frametitle{Kicker name reveal}
<<name_reveal, dependson="informative_missing_data", fig.width=10, out.width="\\textwidth">>=
ggplot(p_d, aes(dist, probability, color=Surface)) +
#  geom_point(data=fg_dist2, aes(dist, p, size=n)) +
  geom_line(aes(group=Surface)) +
  facet_grid(kicker_name ~.) +
  theme_bw() +
  labs(x="Distance (yards)",y="Probability of making the kick", size="Number of\nattempts") +
  ylim(0,1) +
  scale_color_manual(values = c("Grass" = "seagreen", "Synthetic" = "Black"))
@

>>>>

\documentclass[handout]{beamer}

@@

Let's compare two field goal kickers and their proportion of field goals made 
for the 2009-2011 seasons

@@

\frametitle{Raw statistics}
<<data, results="asis">>=
kicker_names <- NFLKicker::kicker_names
kicker_names$kicker_hidden <- c("Kicker 2","Kicker 1")

fg <- NFLKicker::FGdata %>%
  filter(seas >= 2009, gid >= 2388) %>%
  right_join(kicker_names, by = "fkicker")
@

<<raw_statistics, dependson="data", results="asis">>=
fg_raw <- fg %>%
  group_by(kicker_hidden) %>%
  summarize(n = length(good),
            made = sum(good)) %>%
  mutate(p = made/n)

@@

<<proportion_table, dependson="raw_statistics", results="asis">>=
fg_table %>%
  dplyr::select(`Kicker`, `Proportion made`) %>%
  xtable(digits = c(1,NA,2)) %>%
  print(include.rownames = FALSE)
@

@@

%' \begin{frame}
%' \frametitle{Posterior distribution for true probability}
%' 
%' <<posterior, dependson="raw_statistics", fig.width=8>>=
%' create_posterior <- function(d) {
%'   data.frame(probability = seq(0.6,1, length=101)) %>%
%'     mutate(density = dbeta(probability, d$made+.5, d$n-d$made+.5))
%' }
%' 
%' posterior <- fg_raw %>%
%'   group_by(kicker_hidden) %>%
%'   do(create_posterior(.))
%' 
%' ggplot(posterior, aes(probability, density, color=kicker_hidden, group=kicker_hidden)) +
%'   geom_line() +
%'   theme_bw() + 
%'   labs("Probability of making a field goal", ylab="Posterior density", color="Kicker")
%' @
%' \end{frame}

@@

<<fg_dist, dependson="data", fig.width=10, out.width="\\textwidth">>=
fg_dist <- fg %>%
  group_by(kicker_hidden, dist) %>%
  summarize(n = length(good),
            made = sum(good)) %>%
  mutate(p = made/n)

@@

<<number_of_observations, dependson="data", results='asis'>>=
fg_probit <- fg %>%
  mutate(Surface       = ifelse(surf == "Grass", "Grass", "Synthetic"))

@@

<<glm, dependson="number_of_observations", fig.width=10, out.width="\\textwidth">>=
glm_predictions <- function(d) {
  m <- glm(good ~ dist+Surface, data=d, family="binomial")
  dd <- expand.grid(dist=17:80, Surface=c("Grass","Synthetic"))
  dd$probability <- as.numeric(predict(m, dd, type="response"))
  return(dd)
}

@@

\begin{frame}
\frametitle{Hierarchical probit regression model with interaction}
<<hierarchical_model, warning = FALSE>>=
fg2 <- NFLKicker::FGdata %>%
  mutate(Surface       = ifelse(surf == "Grass", "Grass", "Synthetic"),
         scaled_dist = as.numeric(scale(dist, center=36, scale = 10)))

m_glmer <- glmer(good ~ (Surface*scaled_dist|fkicker), 
           data = fg2, 
           family = binomial(link="probit"))
@

@@

For \alert{opportunity} $i$ for kicker $k$, \pause let 
\begin{itemize}
\item $A_{ik}$ be an indicator of attempt, i.e. 1 if a kick was attempted and 0 otherwise, and 
\item $w_{ik}$ be a vector of explanatory variables.
\end{itemize}

@@

The full model is 
\[ \begin{array}{rlrlrl}
Y_{ak} &\ind Ber(\theta_{ak}) & \theta_{ak}  &= \Phi\left(     x_{ak}^\top \beta_k \right) & \beta_k \ind N(\mu,\Sigma) \\
A_{ik} &\ind Ber(\pi_{ik})    & \pi_{ik}     &= \Phi\left(\omega_{ik}^\top \alpha\phantom{_k}  \right) \pause \\ \\
\omega_{ik} &\multicolumn{5}{l}{= (w_{ik}, w_{ik}\Phi^{-1}(\theta_{ik})) = (w_{ik}, w_{ik}x_{ak}^\top \beta_k)}
\end{array} \]
\pause
where 
\begin{itemize}
\item $x_{ak}$ is a set of explanatory variables that affect the probability of making a kick \pause and
\item $w_{ik}$ is a set of explanatory variables that affect the probability of taking a kick. 
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Informative missingness analysis}
<<informative_missing_data, dependson="data">>=
library(dplyr)

@@

<<informative_missing_plot, dependson=c("glm","informative_missing_data"), fig.width=10, out.width="\\textwidth">>=
# fg_dist3 <- fg_probit %>%
#   filter(seas >= 2009, gid >= 2388) %>%
#   group_by(kicker_hidden, dist, Surface) %>%
#   summarize(n = length(good),
#             made = sum(good)) %>%
#   mutate(p = made/n)

ggplot(p_d, aes(dist, probability, color=Surface)) +
  geom_point(data=fg_dist2, aes(dist, p, size=n)) +
  geom_line(aes(group=Surface)) +
  facet_grid(kicker_hidden ~.) +
  theme_bw() +
  labs(x="Distance (yards)",y="Probability of making the kick", size="Number of\nattempts") +
  ylim(0,1) +
  scale_color_manual(values = c("Grass" = "seagreen", "Synthetic" = "Black"))
@
\end{frame}

\begin{frame}
\frametitle{Kicker comparison}

<<kicker_comparison, dependson="informative_missing_data", fig.width=10, out.width="\\textwidth">>=
ggplot(p_d, aes(dist, probability, color=Surface, group=kicker_hidden, linetype=kicker_hidden)) +
  geom_line() +
  facet_grid(Surface ~.) +
  theme_bw() +
  labs(x="Distance (yards)",y="Probability of making the kick", size="Number of\nattempts",
       linetype = "Kicker") +
  ylim(0,1) +
  scale_color_manual(values = c("Grass" = "seagreen", "Synthetic" = "Black"))
@

@@

\begin{frame}
\frametitle{Kicker name reveal}
<<name_reveal, dependson=c("glm","informative_missing_data"), fig.width=10, out.width="\\textwidth">>=
ggplot(p_d, aes(dist, probability, color=Surface, group=kicker_name, linetype=kicker_name)) +
  geom_line() +
  facet_grid(Surface ~.) +
  theme_bw() +
  labs(x="Distance (yards)",y="Probability of making the kick", size="Number of\nattempts",
       linetype = "Kicker") +
  ylim(0,1) +
  scale_color_manual(values = c("Grass" = "seagreen", "Synthetic" = "Black"))
@
